Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array

### 2023-07-31 docs/manuscript/paper.tex
misc text updates

@@

A predictive, first-principles theory of timing noise does not exist at present; there are several plausible physical mechanisms, referenced in Section \ref{sec:intro}. We therefore rely on an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, small-amplitude excursions around a smooth, secular trend. In the model, $f_{\rm p}^{(n)}(t)$ evolves according to the sum of a deterministic torque and a stochastic torque. The deterministic torque is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$. Most PTAs involve millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and the deterministic evolution can be approximated accurately by 
\begin{equation}
 f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
\end{equation} where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. The stochastic torque is assumed to be a zero-mean white noise process. Specifically, the frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas}
\begin{equation}
	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)} +\xi^{(n)}(t) \ . 
	\label{eq:frequency_evolution}
\end{equation}
In Eq \ref{eq:frequency_evolution} $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma^{(n)}$ a proportionality constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ satisfies:
\begin{equation}
	\langle \xi^{(n)}(t) \rangle = 0 \ ,
\end{equation}
\begin{equation}
	\langle \xi^{(n)}(t) \xi^{(n)}(t') \rangle = [\sigma^{(n)}]^2 \delta(t - t') \ .
	\label{eq:xieqn}
\end{equation}
$[\sigma^(n)]^2$ is the variance of $\xi^{(n)}$ and parametrizes the amplitude of the noise, and combined with the mean reversion it gives a characteristic root mean square fluctuations $\approx [\sigma^{(n)}]^2 / \gamma^{(n)}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}. It is important to note that the white noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Eq. \ref{eq:spinevol}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments. \newline

@@

In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave. They are described by the geometric object $\vec{p}$ which we can identify  as the 4-momentum of the radio photon. The presence of a GW induces a shift in the temporal component of the covariant 4-momentum between the emitter and the observer, i.e. $\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $, as \citep[e.g.][]{Maggiore}
\begin{equation}
 \Delta p_t = \frac{\omega}{2} \frac{ h_{ij} (t; \boldsymbol{x}= 0)q^i q^j }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}) }  \left(1 -e^{i \Omega (1 + \boldsymbol{n}\cdot \boldsymbol{q})  d}\right)
	\label{eq:momentum_shift}
\end{equation}
where $\omega$ is the angular pulse frequency $(= 2 \pi f_{\rm p})$ measured in the momentarily comoving reference frame (MCRF) of an observer, $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar. We take the pulsar location $\boldsymbol{q}$ to be constant i.e.  $\boldsymbol{q}$ is not a function of time. In practice the pulsar locations vary with respect to the Earth but are constant with respect to the SSB. This barycentering correction is typically applied when generating TOAs using e.g. {\sc tempo2} and related timing software. Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\vec{u}$ is given by the coordinate-independent expression:
\begin{equation}
	f = p_{\alpha} u^{\alpha} \ . 
	\label{eq:freq_temporal}
\end{equation}
Due to the kinematical corrections from the barycentering process, 
\begin{equation}
	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0)
\end{equation}
where the perturbations to $\vec{u}$ from the GW are of higher order and neglected. In this case the shift in the momentum from Equation \ref{eq:momentum_shift} can be related to a shift in the frequency as,
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}(t)
	\label{eq:measurement}
\end{equation}
where \textcolor{red}{TK: what is the best notation here to give $q^i$ an $(n)$ superscript?}
\begin{equation}
	g^{(n)}(t) = 1 -  \frac{1}{2} \frac{h_{ij} (t; \boldsymbol{x}= 0)q^{i}_{(n)} q^{j}_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \left(1 -e^{i \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}}\right)
	\label{eq:g_func}
\end{equation}
It is will also prove instructive to express Eq. \ref{eq:g_func} in a trigonometric form as,
or in a trigonometric form 
\begin{align}
	g^{(n)}(t) &= 1 - \frac{1}{2} \frac{ H_{ij}q^i_{(n)} q^j_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) } \nonumber \\
	& \times \left[ \cos(-\Omega t +\Phi_0) - \cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d \right) \right]
	\label{eq:g_func_trig}
\end{align}
Equations \ref{eq:measurement}, \ref{eq:g_func_trig}, \ref{eq:g_func}  define a measurement equation that relates the intrinsic pulsar spin frequency to the radio pulse frequency measured by an observer on Earth.

\section{Signal tracking and parameter estimation} \label{sec:detect}
The set of static parameters $\bar{\theta}$ of the model in Section \ref{sec:model} cane be separated into parameters which correspond to the intrinsic frequency evolution of the pulsar and parameters of the GW source, i.e. 
\begin{equation}
	\bar{\theta} =  \bar{\theta}_{\rm psr} \cup \bar{\theta}_{\rm gw}
\end{equation}
with
\begin{equation}
	\bar{\theta}_{\rm psr} = \left \{ \gamma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)},\sigma^{(n)}\right\}_{1\leq n \leq N}
\end{equation}
and
\begin{equation}
	\bar{\theta}_{\rm gw} = \left \{h_{+}, h_{\times}, \delta, \alpha, \psi, \Omega, \Phi_0 \right \} \ . 
\end{equation}
We can reparameterize the two GW polarisation strains in terms of the GW amplitude $h_0$ and the system inclination $\iota$
\begin{align}
	h_+ &= h(1 + \cos^2 \iota) \\
	h_{\times} &= -2h\cos \iota
\end{align}
where $\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$ and the observer line of sight, i.e. $\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$. Henceforth we use the parametrisation in terms of $h_0$ and $\iota$. For a PTA dataset containing $N$ pulsars we have $7 + 5N$ parameters to estimate. Typically the pulsar parameters are better constrained than the GW parameters: for example estimates of pulsar distances are accurate to $\sim$ 10$\%$, but we have no prior information on the source location \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}.The variance of the measurement noise $\Sigma$ can be estimated approximately from the root-mean-square amplitude of phase residuals and we do not treat it as a parameter to be inferred in this work. \newline

In this section we present a new method to infer $\bar{\theta}$ and calculate the marginal likelihood (i.e. the model evidence). In Section \ref{sec:kalman_filter} we outline how noisy measurements of the pulsar frequency $f_{\rm m}^{(n)}(t)$ can be used to estimate the hidden state sequence $f_{\rm p}^{(n)}(t)$ using a Kalman filter. In Section \ref{sec:nested_sampling} we demonstrate how to deploy the Kalman filter in conjunction with a nested sampler to perform Bayesian inference of the model parameters and calculate the model evidence. Model selection and the specification of the null model is described in Section \ref{sec:model_selection}.

\subsection{Kalman filter}\label{sec:kalman_filter}
\begin{figure*}
	%\centering % Not needed
	\begin{subfigure}[b]{1\columnwidth}
		\includegraphics[width=\textwidth]{images/KF1}
		\caption{Using correct parameters}
		\label{fig:6MB_BFS}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{1\columnwidth}
		\includegraphics[width=\textwidth]{images/KF2}
		\caption{Using incorrect parameters}
		\label{fig:25MB_bfs}
	\end{subfigure}
	\caption{Application of a Kalman filter to recover the intrinsic frequency states (top panels) given a measured frequency signal (bottom panels) which has been modulated by the presence of a gravitational wave. In the case where the parameters of the filter are accurate (subfigure (a)) the underlying states are also recovered accurately. Conversely, when the parameters fed to the filter are inaccurate (subfigure (b)) the filter cannot recover the state. \textcolor{red}{TK: this needs to be just one stacked figure.}}
	\label{fig:four figures}
\end{figure*}

@@

The full set of Kalman recursions is presented in Appendix \ref{sec:kalman}. At each discrete timestep $i = 1, ... , M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i$, and the covariance of those estimates $\boldsymbol{P}_i$. The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by projecting the state predictions back into measurement space, $\hat{\boldsymbol{X}}_i \mapsto \hat{\boldsymbol{Y}}_i$, via the measurement equation. The measurement predictions can then be compared aganst the true observed noisy measurements to get a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, sometimes termed the ``innovation". The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{\intercal} \rangle$. The Gaussian log-likelihood can then be calculated at each timestep,
\begin{eqnarray}
	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right )
\end{eqnarray}
with the total log-likelihood the sum over all timesteps, i.e. 
\begin{eqnarray}
	\log \mathcal{L} =  \sum_{i=1}^{i=M} \log \mathcal{L}_i  \label{eq:likelihood}
\end{eqnarray}
The likelihood $\mathcal{L}$, along $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are functions of the static parameters of the model, $\bar{\theta}$. If the estimates of the system parameters, $\hat{\theta}$, that we pass to the Kalman filter are close to the true underlying parameters then the error in $\hat{\boldsymbol{X}}$, $\hat{\boldsymbol{Y}}$ is minimized and $\mathcal{L}$ is maximised. This is illustrated in Figure \ref{fig:6MB_BFS}; given a data timeseries of the measured pulsar frequency the Kalman filter is able to recover the hidden state evolution with high fidelity. Conversely, if $\hat{\theta}$ is not close to $\bar{\theta}$ then the filter is unable to recover the state evolution. This is demonstrated in Figure \ref{fig:25MB_bfs} where the failure of the filter to track the hidden state is evident.

@@

We can use the likelihood returned by the Kalman filter, Eq \ref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\bar{\theta}$ by Bayes' Rule,
\begin{equation}
	p(\bar{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L} \cdot \pi(\theta)}{\mathcal{Z}}
\end{equation}
where $\pi(\theta)$ is the prior distribution on $\bar{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int \mathcal{L} \pi (\theta) d \theta \label{eq:model_evidence}
\end{equation}
In order to estimate the posterior distribution and the model evidence we use nested sampling \citep[NS,][]{Skilling}. Nested sampling is an integration algorithm used for evaluating marginalised likelihood integrals, of the form given by Eq. \ref{eq:model_evidence}, that also returns samples from the posterior $p(\bar{\theta} | \boldsymbol{Y})$. The primary advantage of nested sampling is the ability to compute this evidence integral, which is key for model selection, and proves difficult without considerable extra cost for the usual Markov Chain Monte Carlo (MCMC) approaches. Nested sampling is also typically less computationally intensive than MCMC and can handle multi-modal problems \citep{Ashton2022}. For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, but has also commonly been applied in astrophysics \citep{UltraNest2021}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}. Within this work we exclusively use nested sampling for Bayesian parameter estimation and model selection. \newline

\noindent Multiple nested sampling libraries exist. For gravitational astrophysics it is common to use the dynesty sampler CITE, via the Bilby gravitational wave inference library and we continue to follow this precedent. 
\textcolor{red}{TK: update text here from other version}

\subsection{Model selection}\label{sec:model_selection}
The evidence integral returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a particular model $\mathcal{M}$. This enables us to compare competing models via a Bayes factor,
\begin{equation}
	\beta = \frac{\mathcal{Z}(\mathcal{M}_1)}{\mathcal{Z}(\mathcal{M}_0)}
\end{equation}
where the subscript index identifies a particular model. Throughout this work we take $\mathcal{M}_1$ to be the complete model defined in Section \ref{sec:model}. $\mathcal{M}_0$ is our null model that assumes there is no GW in the data. This is equivalent to setting $g^{(n)}(t)$ from Equation \ref{eq:g_func}, \ref{eq:g_func_trig} $= 1$. The Bayes factors we quote in this paper therefore quantify whether the data shows evidence for a GW compared to no GW.

\section{Tests with synthetic data} \label{sec:testing}

We go on to discuss our choice of pulsars to make up our PTA in Section \ref{sec:pta_pulsars}, before deploying these techniques in Sections \ref{sec:param_estimation} , \ref{sec:detection} for parameter estimation and model selection.

\subsection{Practical considerations}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{booktabs}
\usepackage{multirow}

@@

A predictive, first-principles theory of timing noise does not exist at present; there are several plausible physical mechanisms, referenced in Section \ref{sec:intro}. We therefore rely on an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, small-amplitude excursions around a smooth, secular trend. In the model, $f_{\rm p}^{(n)}(t)$ evolves according to the sum of a deterministic torque and a stochastic torque. The deterministic torque is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$. Most PTAs involve millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and the deterministic evolution can be approximated accurately by 
\begin{equation}
 f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
\end{equation} where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. The stochastic torque is assumed to be a zero-mean white noise process. Specifically, the frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas}
\begin{equation}
	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)} +\xi^{(n)}(t) \ . 
	\label{eq:frequency_evolution}
\end{equation}
In Eq \ref{eq:frequency_evolution}, $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma^{(n)}$ a proportionality constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ satisfies:
\begin{equation}
	\langle \xi^{(n)}(t) \rangle = 0 \ ,
\end{equation}
\begin{equation}
	\langle \xi^{(n)}(t) \xi^{(n)}(t') \rangle = [\sigma^{(n)}]^2 \delta(t - t') \ .
	\label{eq:xieqn}
\end{equation}
$[\sigma^(n)]^2$ is the variance of $\xi^{(n)}$ and parametrizes the amplitude of the noise, and combined with the mean reversion it gives a characteristic root mean square fluctuations $\approx [\sigma^{(n)}]^2 / \gamma^{(n)}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}. It is important to note that the white noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Eq. \ref{eq:spinevol}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments. \newline

@@

In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave. They are described by the geometric object $\vec{p}$ which we can identify  as the 4-momentum of the radio photon. The presence of a GW induces a shift in the temporal component of the covariant 4-momentum between the emitter and the observer, i.e. $\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $, as \citep[e.g.][]{Maggiore}
\begin{equation}
 \Delta p_t = \frac{\omega}{2} \frac{ h_{ij} (t; \boldsymbol{x}= 0)q^i q^j }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}) }  \left(1 -e^{i \Omega (1 + \boldsymbol{n}\cdot \boldsymbol{q})  d}\right)
	\label{eq:momentum_shift}
\end{equation}
where $\omega$ is the angular pulse frequency $(= 2 \pi f_{\rm p})$ measured in the momentarily comoving reference frame (MCRF) of an observer, $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar. We take the pulsar location to be constant i.e.  $\boldsymbol{q}$ is not a function of time. In practice the pulsar locations vary with respect to the Earth but are constant with respect to the SSB. This barycentering correction is typically applied when generating TOAs using e.g. {\sc tempo2} and related timing software. Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\vec{u}$ is given by the coordinate-independent expression:
\begin{equation}
	f = p_{\alpha} u^{\alpha} \ . 
	\label{eq:freq_temporal}
\end{equation}
Due to the kinematical corrections from the barycentering process, 
\begin{equation}
	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0)
\end{equation}
where the perturbations to $\vec{u}$ from the GW are of higher order and neglected. In this case the shift in the momentum from Equation \ref{eq:momentum_shift} can be related to a shift in the frequency as,
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}(t)
	\label{eq:measurement}
\end{equation}
where \textcolor{red}{TK: what is the best notation here to give $q^i$ an $(n)$ superscript?}
\begin{equation}
	g^{(n)}(t) = 1 -  \frac{1}{2} \frac{h_{ij} (t; \boldsymbol{x}= 0)q^{i}_{(n)} q^{j}_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \left(1 -e^{i \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}}\right)
	\label{eq:g_func}
\end{equation}
It is will also prove instructive to express Eq. \ref{eq:g_func} in a trigonometric form as,
or in a trigonometric form 
\begin{align}
	g^{(n)}(t) &= 1 - \frac{1}{2} \frac{ H_{ij}q^i_{(n)} q^j_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) } \nonumber \\
	& \times \left[ \cos(-\Omega t +\Phi_0) - \cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d \right) \right]
	\label{eq:g_func_trig}
\end{align}
Equations \ref{eq:measurement}, \ref{eq:g_func_trig}, \ref{eq:g_func}  define a measurement equation that relates the intrinsic pulsar spin frequency to the radio pulse frequency measured by an observer on Earth.

\section{Signal tracking and parameter estimation} \label{sec:detect}
The set of static parameters $\boldsymbol{\theta}$ of the model in Section \ref{sec:model} can be separated into parameters which correspond to the intrinsic frequency evolution of the pulsar and parameters of the GW source, i.e. 
\begin{equation}
	\boldsymbol{\theta} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta}_{\rm gw} \label{eq:params1}
\end{equation}
with
\begin{equation}
	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)},\sigma^{(n)}\right\}_{1\leq n \leq N}
\end{equation}
and
\begin{equation}
	\boldsymbol{\theta}_{\rm gw} = \left \{h_{+}, h_{\times}, \delta, \alpha, \psi, \Omega, \Phi_0 \right \} \ .  \label{eq:params3}
\end{equation}
We can reparameterize the two GW polarisation strains in terms of the GW amplitude $h_0$ and the system inclination $\iota$
\begin{align}
	h_+ &= h(1 + \cos^2 \iota) \\
	h_{\times} &= -2h\cos \iota
\end{align}
where $\iota$ is the angle between the normal to the SMBHB orbital plane, $\boldsymbol{L}$ and the observer line of sight, i.e. $\cos \iota = \boldsymbol{n} \cdot \boldsymbol{L}$. Henceforth we use the parametrisation in terms of $h_0$ and $\iota$. For a PTA dataset containing $N$ pulsars we have $7 + 5N$ parameters to estimate. Typically the pulsar parameters are better constrained than the GW parameters: for example estimates of pulsar distances are accurate to $\sim$ 10$\%$, but we have no prior information on the source location \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}.The variance of the measurement noise $\Sigma$ can be estimated approximately from the root-mean-square amplitude of phase residuals and we do not treat it as a parameter to be inferred in this work. \newline

In this section we present a new method to infer $\boldsymbol{\theta}$ and calculate the marginal likelihood (i.e. the model evidence). In Section \ref{sec:kalman_filter} we outline how noisy measurements of the pulsar frequency $f_{\rm m}^{(n)}(t)$ can be used to estimate the hidden state sequence $f_{\rm p}^{(n)}(t)$ using a Kalman filter. In Section \ref{sec:nested_sampling} we demonstrate how to deploy the Kalman filter in conjunction with a nested sampler to perform Bayesian inference of the model parameters and calculate the model evidence. Model selection and the specification of the null model is described in Section \ref{sec:model_selection}.

%The choice of prior distributions on the model parameters are described in

\subsection{Kalman filter}\label{sec:kalman_filter}
\begin{figure*}
	%\centering % Not needed
	\begin{subfigure}[b]{1\textwidth}
		\includegraphics[width=\textwidth]{images/Kalman_example_true_params}
		\caption{Kalman filter using correct estimate of the static parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ }
		\label{fig:6MB_BFS}
	\end{subfigure} \newline

	\begin{subfigure}[b]{1\textwidth}
		\includegraphics[width=\textwidth]{images/Kalman_example_wrong_params}
		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
		\label{fig:25MB_bfs}
	\end{subfigure}
	\caption{Application of a Kalman filter to estimate the intrinsic pulsar frequency state $f_{\rm p}(t)$ given a measured frequency $f_{\rm m}(t)$. We here plot the ephemeris subtracted state $f'_{\rm p}(t) = f_{\rm p}(t) - f_{\rm em} (t)$ to better illustrate the stochastic wandering of the pulsar frequency. In (a) the Kalman filter is run using identically correct estimates of the static system parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$. In (b) the Kalman filter is run with $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$. The top left panel shows the true pulsar state $f'_{\rm p}(t)$ (blue line) and the state estimated by the Kalman filter, $\hat{f'}_{\rm p}(t)$ (green line). The middle left panel shows the true $f_{\rm m}(t)$ (red line) and the measured frequency estimated  by the Kalman filter, $\hat{f}_{\rm m}(t)$ (magenta). The bottom left panel shows the residual or the innovation $\epsilon(t) =f_{\rm m}(t) - \hat{f}_{\rm m}(t)$. The right hand panel plots the distribution of the innovation over the total observation period.}
	\label{fig:four figures}
\end{figure*}

@@

The full set of Kalman recursions is presented in Appendix \ref{sec:kalman}. At each discrete timestep $i = 1, ... , M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i$, and the covariance of those estimates $\boldsymbol{P}_i$. The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by projecting the state predictions back into measurement space, $\hat{\boldsymbol{X}}_i \mapsto \hat{\boldsymbol{Y}}_i$, via the measurement equation, Eq. \ref{eq:measurement}d. The measurement predictions can then be compared aganst the true observed noisy measurements to get a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, sometimes termed the ``innovation". The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{\intercal} \rangle$. The Gaussian log-likelihood can then be calculated at each timestep,
\begin{eqnarray}
	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right )
\end{eqnarray}
with the total log-likelihood the sum over all timesteps, i.e. 
\begin{eqnarray}
	\log \mathcal{L} =  \sum_{i=1}^{i=M} \log \mathcal{L}_i  \label{eq:likelihood}
\end{eqnarray}
The likelihood $\mathcal{L}$, along $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are functions of the static parameters of the model, $\bar{\theta}$. If the estimates of the system parameters, $\hat{\theta}$, that we pass to the Kalman filter are close to the true underlying parameters then the error in $\hat{\boldsymbol{X}}$, $\hat{\boldsymbol{Y}}$ is minimized and $\mathcal{L}$ is maximised. This is illustrated in Figure \ref{fig:6MB_BFS}; given a data timeseries of the measured pulsar frequency the Kalman filter is able to recover the hidden state evolution with high fidelity - leavign only gaussian errrs

Conversely, if $\hat{\theta}$ is not close to $\bar{\theta}$ then the filter is unable to recover the state evolution. This is demonstrated in Figure \ref{fig:25MB_bfs} where the failure of the filter to track the hidden state is evident.

@@

We can use the likelihood returned by the Kalman filter, Eq \ref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\bar{\theta}$ by Bayes' Rule,
\begin{equation}
	p(\bar{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L} \cdot \pi(\theta)}{\mathcal{Z}}
\end{equation}
where $\pi(\theta)$ is the prior distribution on $\bar{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int \mathcal{L} \pi (\theta) d \theta \label{eq:model_evidence}
\end{equation}
In order to estimate the posterior distribution and the model evidence we use nested sampling \citep[NS,][]{Skilling} throughout this work. Nested sampling is an integration algorithm used for evaluating marginalised likelihood integrals, of the form given by Eq. \ref{eq:model_evidence}, that also returns samples from the posterior $p(\bar{\theta} | \boldsymbol{Y})$. The primary advantage of nested sampling is the ability to compute this evidence integral, which is key for model selection, and proves difficult without considerable extra cost for the usual Markov Chain Monte Carlo (MCMC) approaches. Nested sampling is also typically less computationally intensive than MCMC and can handle multi-modal problems \citep{Ashton2022}. For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, but has also commonly been applied in astrophysics \citep{UltraNest2021}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}. For a review of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}. Multiple nested sampling algorithms and computational libraries exist. \citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}. For gravitational astrophysics it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \cite{bilby.507.2037A} front-end library. We  follow this precedent and use \texttt{Bilby} for all Bayesian inference in this work.

%from https://arxiv.org/pdf/2102.12478.pdf
%Nested sampling performs these dual processes by evolving an
%ensemble of 𝑛live live points by initially drawing them from the prior
%distribution 𝜋, and then at each subsequent iteration: (a) discarding
%the lowest likelihood live point and (b) replacing it by a live point
%drawn from the prior 𝜋 subject to hard constraint that the new point is
%at a higher likelihood than the discarded point. The set of discarded
%dead points and likelihoods can be used to compute a statistical
%estimate of the Bayesian evidence (1), and be weighted to form a set
%of samples from the posterior distribution.
%The nested sampling procedure therefore differs substantially from
%traditional strategies such as Gibbs sampling, Metropolis-Hastings
%or Hamiltonian Monte Carlo in that it is acts as a likelihood scanner
%rather than a posterior sampler. It scans athermally, so in fact is
%capable of producing samples at any temperature and computing the
%full partition function in addition to the Bayesian evidence (Handley
%2019b).

\subsection{Model selection}\label{sec:model_selection}
The evidence integral returned by nested sampling is the probability of the data $\boldsymbol{Y}$ given a particular model $\mathcal{M}$. This enables us to compare competing models via a Bayes factor,
\begin{equation}
	\beta = \frac{\mathcal{Z}(\mathcal{M}_1)}{\mathcal{Z}(\mathcal{M}_0)}
\end{equation}
where the subscript index identifies a particular model. Throughout this work we take $\mathcal{M}_1$ to be the complete model defined in Section \ref{sec:model}. $\mathcal{M}_0$ is our null model that assumes there is no GW in the data. This is equivalent to setting $g^{(n)}(t)=1$ in Equation \ref{eq:g_func}, \ref{eq:g_func_trig}. The Bayes factors we quote in this paper therefore quantify whether the data shows evidence for a GW compared to no GW.

\section{Tests with synthetic data} \label{sec:testing}
We go on to discuss our choice of pulsars to make up our PTA in Section \ref{sec:pta_pulsars}, before deploying these techniques in Sections \ref{sec:param_estimation} , \ref{sec:detection} for parameter estimation and model selection.

\subsection{Synthetic data generation}

\subsubsection{Constructing a synthetic PTA}\label{sec:synt_pta}

\begin{figure}
	\includegraphics[width=\columnwidth]{images/pulsar_distribution}
	\caption{Spatial distribution of 47 pulsars from the 12.5 year NANOGrav data release that make up the synthetic PTA used in this work. The pulsar distances relative to the observer are also indicated, with the majority of pulsars having a distance $\sim 1$ kpc. The grey dashed line denotes the Galactic plane.}
	\label{fig:pulsar_distrib}
\end{figure}
